The complex structure and peculiar internal motions of the planetary nebula NGC 6210

### 2020-05-22 ms.tex
Finish ionization discussion, start axis orientation discussion

 \begin{figure}
\centering
  \includegraphics[width=0.4\textwidth]{tere-figs/Figure2a}
  \caption{ }
\end{figure}

\begin{table}
\centering
%\caption{Log  of Time-resolved  Observations of {\bf NGC~6210}.}
\begin{tabular}{l|cccccccc} \hline 
&  \multicolumn{7}{c}{Spectroscopy hight resolution/ 2.1m  OAN-SPM}            \\[0.1pt]
\hline 
Run    &   Exp. & Range  & Slit & P. A.   & slit        \\
Epoch   &     Time        &    (s)   &  (\AA)    &    (\AA)  & (m$\mu$)     \\[1pt]   \hline 
28/06/1998          & 30 & H$\alpha$  & 150 & 90 & O\\
28/06/1998          & 300 & H$\alpha$  & 150 & 90 & N\\
28/06/1998           & 1200 & H$\alpha$   & 150 & 90 & M,P,Q \\
05/06/2003         & 1800 &  H$\alpha$   & 70 &0 &     A,B,D*,E*,F,H,I,K\\ 
16/10/2003   &  1800 &   H$\alpha$    & 70 & $-$21 & T\\
16/10/2003    &  1800 &   H$\alpha$   & 70 & $-$68 & V\\
17/10/2003    &  1800 &   H$\alpha$  & 70 & 77 & W\\
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$9 & R  \\  
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$19 & S  \\  
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$56 & U  \\  
14/06/2004   & 1800 &   O\sc{iii}  & 150 & 0 & A',L  \\
21/05/2011     & 1800 & H$\alpha$  & 150 & 0 & G* \\
21/05/2011     & 600 & O\sc{iii} & 150 & 0  & G \\
06/07/2013    &  1800 & H$\alpha$   & 150 & 0 & C,I,J \\
18/08/2015  &  1800 &   H$\alpha$, [OIII]  & 70 & 0 & C,D,E,F,G\\
19/08/2015  &  1800 &   H$\alpha$, [OIII]   & 70 & 0 & B,A*,I\\
20/08/2015   &  1800 &   H$\alpha$, [OIII] & 70 & 0 & H,J,K\\
18/09/2019  &  1800 &   H$\alpha$    & 150 & 56 & \\

\hline
\end{tabular}
\label{table:pa5}
\end{table}

@@

\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth]{tere-figs/Figure3}
  \caption{}
\end{figure*}

@@

\begin{figure}
\centering
%\includegraphics[width=0.7\textwidth]{tere-figs/Figure8}
%\includegraphics[width=0.327\textwidth]{tere-figs/opo9836f}
%\includegraphics[width=0.32\textwidth]{tere-figs/N6210_ruso}
\includegraphics[width=0.6\textwidth]{sketch}

  \caption{}
\end{figure}

@@

\clearpage

@@

@@

@@

For those nebular features where both proper motion and radial velocity measurements exist,
it is possible to estimate the full three-dimensional velocity vector.
Under the assumption that all motions are strictly radial,
one can also find the three-dimensional position.
These are given in Table~\ref{tab:3d} for a wide variety of nebular features
in the inner shells, intermediate shell, low-ionization knots, and outer lobes.

@@

Note that the uncertainties given in the table are derived from observational errors only,
and they do not include the systematic uncertainty in the distance

@@

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/vel-radius-systems-annotated.pdf}
  \caption{
    True velocity, \(V_{\text{tot}}\), versus true radius, \(R\), for all the features listed in Table~\ref{tab:3d}.
    Diagonal lines show dynamical times between \num{1000} and \SI{16000}{years}, as marked.
    Symbol shape indicates the type of feature:
    circle symbols are high-ionization shells (\S~\ref{sec:high-ioniz-shells}),
    cross symbols are low-ionization knots or knot complexes (\S~\ref{sec:knot-complexes}),
    diamond symbols are high-velocity and intermediate-velocity systems in the outer lobes (\S~\ref{sec:outer-lobes}).
    Error bars show the observational uncertainties given in the table,
    except for outer lobe system \SysM{2}, where the lack of proper motion measurements mean that only lower limits for \(V_{\text{tot}}\) and \(R\) are available, indicated by arrows.
    The color of each symbol corresponds the position angle of the feature, as shown by the key at right,
    which also indicates the values for each of the four axes: A, B, C, and D.
    The position angle is calculated modulo \ang{180} so that the positive and negative arm of each axis have the same color.
    Also shown (gray ellipses) are the expansion velocity and range of radii of the inner and outer halo features (\S~\ref{sec:haloes}).
      }
  \label{fig:ages}
\end{figure}

@@

In a similar way, the ionized mass of each component can be calculated as
\begin{equation}
  \label{eq:mass}
  M = \frac{4\pi D^2\, \bar{m}\, F(\Ha) }{\alpha_{\text{eff}}\, E(\Ha)\, n} \ , 
\end{equation}
where \(D\) is the distance from Earth (assumed \SI{2}{kpc}) and \(\bar{m} \approx \SI{2.17e-24}{g}\) is the mean mass per hydrogen nucleon.
The total ionized mass of the nebula is found to be \SI{0.372}{\msun} and results for individual components are shown in column~4 of Table~\ref{tab:summary}.
It is interesting to note that the high-density components (inner shell plus knot complexes)
only represent about one-third of the total ionized mass,
despite dominating the emission line flux.
The intermediate shell is the most massive single component (\SI{0.135}{\msun}),
also roughly one-third of the total,
with the remaining third corresponding to the outer lobes and halos.

@@

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic age of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shells that have been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, the estimates are only for up to \SI{1500}{yr} ago, the kinematic timescale of the inner shells.
The mass-loss rate is expected to have declined significantly since then as the central star transitions to higher effective temperature and smaller radius,
eventually reaching the present-day value of \(\dot{M} = \SI{9.12e-9}{M_\odot.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.

@@

It is apparent from the table and figure that there is a wide range of ionization parameters in the nebula.
The steep decline in density with radius (faster than \(R^{-2}\)) means that the ionization parameter tends to increase with radius, although the lowest values are associated with the knot complexes, owing to their relatively high densities as compared with other features at similar radii.
Figure~\ref{fig:line-ratios} shows that the \nii/\Ha{} ratio varies by two orders of magnitude and is clearly anti-correlated with \ionpar.
The \oiii/\Hb{} ratio, on the other hand, shows no correlation with \ionpar{} and indeed varies little within the nebula.

>>>>

 \begin{figure}
\centering
  \includegraphics[width=0.4\textwidth]{tere-figs/Figure2a}
  \caption{ }
  \label{fig:slit-positions}
\end{figure}

\begin{table}
\centering
\caption{OAN-SPM/Mezcal high-resolution spectroscopy of NGC~6210}
\label{table:pa5}
\begin{tabular}{l|cccccccc} \toprule
  &   Exp.~Time & Spectral  & Slit & P. A.   & slit        \\
  Epoch   &    (s) & Range  &  (\AA)    &    (\AA)  & (m$\mu$)     \\
  \midrule
1998/06/28          & 30 & \Ha  & 150 & 90 & O\\
1998/06/28          & 300 & \Ha  & 150 & 90 & N\\
1998/06/28           & 1200 & \Ha   & 150 & 90 & M,P,Q \\
2003/06/05         & 1800 &  \Ha   & 70 &0 &     A,B,D*,E*,F,H,I,K\\ 
2003/10/16   &  1800 &   \Ha    & 70 & $-$21 & T\\
2003/10/16    &  1800 &   \Ha   & 70 & $-$68 & V\\
2003/10/17    &  1800 &   \Ha  & 70 & 77 & W\\
2004/06/13   & 1800 &   \oiii  & 70 & $-$9 & R  \\  
2004/06/13   & 1800 &   \oiii  & 70 & $-$19 & S  \\  
2004/06/14    & 1800 &  \sii &150 & $-$19 & S  \\
2004/06/13   & 1800 &   \oiii  & 70 & $-$56 & U  \\  
2004/06/14   & 1800 &   \oiii  & 150 & 0 & A',L  \\
2011/05/21     & 1800 & \Ha  & 150 & 0 & G* \\
2011/05/21     & 600 & \oiii & 150 & 0  & G \\
2013/07/06    &  1800 & \Ha   & 150 & 0 & C,I,J \\
2015/08/18  &  1800 &   \Ha, \oiii  & 70 & 0 & C,D,E,F,G\\
2015/08/19  &  1800 &   \Ha, \oiii   & 70 & 0 & B,A*,I\\
2015/08/20   &  1800 &   \Ha, \oiii & 70 & 0 & H,J,K\\
2019/09/18  &  1800 &   \Ha    & 150 & 56 & \\
  \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth]{tere-figs/Figure3}
  \caption{}
  \label{fig:pv-array}
\end{figure*}

@@

\begin{figure}
\centering
%\includegraphics[width=0.7\textwidth]{tere-figs/Figure8}
%\includegraphics[width=0.327\textwidth]{tere-figs/opo9836f}
%\includegraphics[width=0.32\textwidth]{tere-figs/N6210_ruso}
\includegraphics[width=\linewidth]{sketch}
\caption{}
\label{fig:hst}
\end{figure}

@@

@@

The knot complexes are also visible in the \oiii{} spectra,
although they are fainter than the shells in this line.
There is also a slight difference in velocity,
with \oiii{} showing expansion velocities that are \num{2} to \SI{5}{km.s^{-1}} smaller than \nii{}.
If the expansion velocities within the knot complexes are assumed to increase with radius,
then this is evidence for ionization stratification along the line of sight
with the \oiii{} emission arising slightly closer to the star than \nii{}.
This is also consistent with the HST images (e.g., Figure~\ref{fig:hst}d),
which show that the \oiii{} emission associated with the complexes is more diffuse than the compact \nii{} knots.

@@

    SE knot & 135 &  & -35 \pm 3 & 39 \pm 10 & 57 \pm 7 & -47 \pm 8 & 8 \pm 1 & 0.09 \pm 0.02 & 1.5 \pm 0.4\\

@@

For those nebular features where both proper motion and radial velocity measurements exist,
it is possible to estimate the full three-dimensional velocity vector.
Under the assumption that all motions are strictly radial,
one can also find the three-dimensional position.
These are given in Table~\ref{tab:3d} for a wide variety of nebular features
in the inner shells, intermediate shell, low-ionization knots, and outer lobes.\footnote{
  Note that the uncertainties given in the table are derived from observational errors only,
  and they do not include the systematic uncertainty in the distance.
}

@@

@@

An approximate three-dimensional visualization of the nebular structure is given in Figure~\ref{fig:cut-axis-3d}.
The three axes, A, C, and D, that define the outer lobes all have small inclinations,
\(|i| < \ang{30}\) to the plane of the sky,
while the knot complexes all have higher inclinations,
\(|i| > \ang{45}\).

It is apparent from this figure that the three axes that define the outer lobes,
A, B, and D,
can be grouped into a loose conical structure with a half-opening angle of 
\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/vel-radius-systems-annotated.pdf}
  \caption{
    True velocity, \(V_{\text{tot}}\), versus true radius, \(R\), for all the features listed in Table~\ref{tab:3d}.
    Diagonal lines show dynamical times between \num{1000} and \SI{16000}{years}, as marked.
    Symbol shape indicates the type of feature:
    circle symbols are high-ionization shells (\S~\ref{sec:high-ioniz-shells}),
    cross symbols are low-ionization knots or knot complexes (\S~\ref{sec:knot-complexes}),
    diamond symbols are high-velocity and intermediate-velocity systems in the outer lobes (\S~\ref{sec:outer-lobes}).
    Error bars show the observational uncertainties given in the table,
    except for outer lobe system \SysM{2}, where the lack of proper motion measurements mean that only lower limits for \(V_{\text{tot}}\) and \(R\) are available, indicated by arrows.
    The color of each symbol corresponds the position angle of the feature, as shown by the key at right,
    which also indicates the values for each of the four axes: A, B, C, and D.
    The position angle is calculated modulo \ang{180} so that the positive and negative arm of each axis have the same color.
    Also shown (gray ellipses) are the expansion velocity and range of radii of the inner and outer halo features (\S~\ref{sec:haloes}).
      }
  \label{fig:ages}
\end{figure}

@@

Once the density is known, then the column density of ionized hydrogen follows as proportional to \(S(\Ha)/n\) and 
the ionized mass of each component can be calculated as
\begin{equation}
  \label{eq:mass}
  M = \frac{4\pi D^2\, \bar{m}\, F(\Ha) }{\alpha_{\text{eff}}\, E(\Ha)\, n} \ , 
\end{equation}
where \(D\) is the distance from Earth (assumed \SI{2}{kpc}) and \(\bar{m} \approx \SI{2.17e-24}{g}\) is the mean mass per hydrogen nucleon.
The total ionized mass of the nebula is found to be \SI{0.372}{\msun} and results for individual components are shown in column~4 of Table~\ref{tab:summary}.
It is interesting to note that the high-density components (inner shell plus knot complexes)
only represent about one-third of the total ionized mass,
despite dominating the emission line flux.
The intermediate shell is the most massive single component (\SI{0.135}{\msun}),
also roughly one-third of the total,
with the remaining third corresponding to the outer lobes and halos.

@@

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic age of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shells that have been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, this method can only estimate mass loss before \SI{1500}{yr} ago, which is the kinematic timescale of the inner shells.
Since that time, the mass-loss rate is expected to have declined significantly
due to the central star transitioning to higher effective temperature and smaller radius,
eventually reaching the present-day value of \(\dot{M} = \SI{9.12e-9}{M_\odot.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.
This is indicated on the figure by the gray downward pointing arrow.

@@

It is apparent from the table and figure that there is a wide range of ionization parameters in the nebula.
The steep decline in density with radius (faster than \(R^{-2}\)) means that the ionization parameter tends to increase with radius, although the lowest values are associated with the knot complexes, owing to their relatively high densities as compared with other features at similar radii.
Figure~\ref{fig:line-ratios} shows that the \nii/\Ha{} ratio varies by two orders of magnitude and is clearly anti-correlated with \ionpar.
The \oiii/\Hb{} ratio, on the other hand, shows no correlation with \ionpar{} and indeed varies little within the nebula.
This is what is expected in an optically thin nebula, where the doubly ionized metals predominate.

As an example, gray solid lines in Figure~\ref{fig:line-ratios} show a toy photoionization model,
in which we assume
\(
\chem{O^{2+}} / \chem{O^+} = \chem{N^{2+}} / \chem{N^+} = A \, \ionpar ,
\)
where \(A \approx \num{1.6e4}\) is a dimensionless constant,
whose value we fix by requiring the model to approximately fit the \(\nii/\Ha\) observations.\footnote{
  Setting the \chem{N} and \chem{O} ion fractions to be equal in this equation is tantamount to assuming an \chem{N^+/O^+} ionization correction factor (ICF) of unity \citep{Kingsburgh:1994a}.
  Fits to a large suite of photoionization models
  \citetext{equations~[14--16] of \citealp{Delgado-Inglada:2014b}}
  imply \(\chem{ICF(N^+/O^+)} = 1.4 \pm 1.0\) for the nebular parameters of NGC~6210,
  whereas spectrophotometry of the whole nebula \citep[Table~7]{Pottasch:2009a} yields  \(\chem{ICF(N^+/O^+)} = 0.78\).
  Both of these estimates are sufficiently close to unity for our purposes.
  % In principle, the value of \(A\) may be slightly different for \chem{O} and \chem{N}.
  % It can be calculated as \(A = f_+ \sigma_+ c / \alpha_+\),
  % where \(f_+\) is the fraction of H-ionizing photons that are also capable of ionizing the singly ionized metal
  % (\(h\nu > \SI{29.6}{eV}\) for \chem{N^+} or \SI{35.1}{eV} for \chem{O^+})
  % and \(\sigma_+\), \(\alpha_+\) are respectively the effective photoionization cross-section and recombination rate.
  % A simple estimate using data from \citet[Appendix~A5]{Osterbrock:2006a}
}
In the model, we calculate the line emissivities using \textsc{Pyneb},
assuming electron temperatures and abundances as marked on the figures,
which are based on the values derived by \citet{Pottasch:2009a}.
Selected ionic fractions are indicated by white dots, which show that,
even in the \nii{}-brightest knots,
the \chem{N^+/N} ratio never exceeds \num{0.1}.

One caveat here is that the measured line ratios correspond to the integrated emission along the line of sight, whereas, 
in the case of the knot complexes,
there is spectroscopic evidence for ionization stratification
(see final paragraph of \S~\ref{sec:knot-complexes}),
which means that the \nii{} and \oiii{} emission may come from distinct zones.%
\footnote{
  The \Ha{} and \Hb{} emission would come from both zones but the \oiii{} zone would predominate due to its larger emission measure.
}
It is therefore possible that \emph{locally} \chem{N^+/N} may reach unity,
but that the effect is masked by integration along the line of sight.

Although we have demonstrated that the nebula as a whole is optically thin in the \chem{H^0}- and \chem{He^0}-ionizing continua,
this is clearly not the case for the \chem{He^+}-ionizing continuum (\(h\nu > \SI{54.4}{eV}\)),
as evidenced by the compact distribution of \heii{} emission at the inner edge of the high-ionization shells (see Figs.~\ref{fig:heii-shell-annotated}, \ref{fig:heii-shell-components}, and \ref{fig:heii-shell-velocity-axes}).
This is consistent with the presence of an ionization-bounded \chem{He^{2+}} Strömgren sphere,
with radius \(\approx \SI{0.04}{pc}\),
which traps all the \chem{He^+}-ionizing radiation.
An interesting consequence is that,
even though the ionization parameter increases with distance from the central star,
there is likely to be no \chem{O^{3+}} in the outer nebula since the photoionization of \chem{O^{2+}} requires \(h\nu > \SI{54.9}{eV}\).
Indeed, our measurements (Fig.~\ref{fig:line-ratios}b) show that \oiii{}/\Hb{} remains high in the lobes and halo,
consistent with \chem{O^{2+}/O \approx 1}.
